Theory about Normal Basis on $F(2^k)$

### 2017-03-03 dissertation/submission/prelim_GF.tex
after submission: revision 2

\chapter{Finite Fields and Sequential Arithmetic Circuits}
\label{ch:prelim_GF}
This chapter provides a mathematical background for understanding 
finite fields (Galois fields) and explains how to design Galois field (GF) arithmetic circuits.
We first introduce the mathematical concepts of groups, rings, fields, and 
polynomials. 
We then apply these concepts to create Galois field arithmetic functions and 
explain how to map them to a Boolean circuit implementation.
Additionally, we introduce a special type of sequential arithmetic hardware based on normal basis, as well
as the normal basis theory behind the designing such hardware.

The material is referred from \cite{galois_field:mceliece, ftheory:2006, ff:1997} for Galois field concepts, 
\cite{mastro:1989, PT:1985, acar:1998, wu:2002, Knezevic:2008} for hardware design over Galois fields 
and previous work by Lv \cite{lv:phd}.
Normal basis theory in this section is referred from \cite{normal_book, gao:phd_normal_basis} and sequential
normal basis arithmetic hardware designs come from \cite{mullinONB,MasseyOmura,agnew1991implementation, RHmulti}.

@@

\begin{Definition}
A {\bf field} $\mathbb{F}$ is a commutative ring with unity, where every
non-zero element in $\mathbb{F}$ has a multiplicative inverse; {\it i.e.} $\forall
a \in \mathbb{F} - \{0\}$, $\exists \hat{a} \in \mathbb{F}$ such that $ a \cdot
\hat{a} = 1$.
\end{Definition}

A field is defined as a ring with one extra condition: the presence of a 
multiplicative inverse for all non-zero elements.
Therefore, a field must be a ring while a ring is not necessarily a field.
For example, the set $\mathbb{Z}_{2^k} = \{0,1,\cdots, 2^k-1\}$ forms a finite ring.
However, $\mathbb{Z}_{2^k}$ is not a field because not every element in
$\mathbb{Z}_{2^k}$ has a multiplicative inverse. 
In the ring $\mathbb{Z}_{2^3}$, for 
instance, the element $5$ has an inverse ($5\cdot5\pmod{8}=1$) but the element $4$
does not.

An important concept in field theory is {\bf field extension}. The idea behind a
field extension is to take a base field and construct a larger field which 
contains the base field as well as satisfies additional properties. For example,
the set of real numbers $\mathbb{R}$ forms a field; one extension of 
$\mathbb{R}$ is the set of complex numbers $\mathbb{C}=\mathbb{R}(i)$. Every
element of $\mathbb{C}$ can be represented as $a+b\cdot i$ where $a,b \in \mathbb{R}$,
hence $\mathbb{C}$ is a two-dimensional extension of $\mathbb{R}$.

@@

\subsection{Finite Field}
Finite fields find widespread applications in 
many areas of electrical engineering and computer science such as error-
correcting codes, elliptic curve cryptography, digital signal processing, 
testing of VLSI circuits, among others.
In this dissertation, we specifically focus on their application to 
the FSM traversal of sequential Galois field circuits as well as abstraction refinement based on UNSAT core extraction.
This section describes the relevant Galois field concepts
\cite{galois_field:mceliece} \cite{ftheory:2006} \cite{ff:1997}
and hardware arithmetic designs over such fields \cite{mastro:1989} \cite{PT:1985} 
\cite{acar:1998} \cite{wu:2002} \cite{Knezevic:2008}.

@@

Algebraic extensions of the binary field $\F_{2}$  
are generally termed as {\it binary extension fields} $\Fkk$.
Where elements in $\F_2$ can only represent $1$ bit, elements in $\Fkk$ 
represent a $k$-bit vector.
This allows them to be widely used in digital hardware applications.
In order to construct a Galois field of the form $\Fkk$, 
an {\bf irreducible polynomial} is required:
\begin{Definition}
A polynomial $P(x) \in \mathbb{F}_{2}\left[x\right]$, {\it i.e.} the set of all 
polynomials in $x$ with coefficients in $\F_2$, is {\bf irreducible} 
if $P(x)$ is non-constant with degree $k$ and cannot be 
factored into a product of polynomials of lower degree in $\mathbb{F}_2[x]$.
\end{Definition}

Therefore, the polynomial $P(x)$ with degree $k$ is irreducible over 
$\mathbb{F}_{2}$ if and only if it has no roots in $\mathbb{F}_{2}$,
i.e if $\forall a \in \mathbb{F}_{2}$, $P(a)\neq 0$.
For example, $x^2+x+1$ is an irreducible polynomial over $\mathbb{F}_{2}$
because it has no solutions in $\mathbb{F}_{2}$, {\it i.e.} $(0)^2+(0)+1=1\neq0$ 
and $(1)^2+(1)+1=1\neq0$ over $\F_2$.
Irreducible polynomials exist for any degree $\geq 2$ in $\mathbb{F}_2[x]$.

Given an irreducible polynomial $P(x)$ of degree $k$ in the polynomial ring 
$\mathbb{F}_2[x]$, we can construct a binary extension field 
$\mathbb{F}_{2^k} \equiv \mathbb{F}_2[x] \pmod{P(x)}$.
Let $\alpha$ be a root of $P(x)$, {\it i.e.} $P(\alpha)=0$.
Since $P(x)$ is irreducible over
$\mathbb{F}_2[x]$, $\alpha \notin \mathbb{F}_2$. 
Instead, $\alpha$ is an element in $\mathbb{F}_{2^k}$. 
Any element $A \in \mathbb{F}_{2^k}$ is then represented as: 
\begin{equation}\label{rep:poly}
A= \sum_{i=0}^{k-1} (a_i \cdot \alpha^i) = a_0 + a_1\cdot\alpha + \cdots + a_{k-1}\cdot \alpha^{k-1}\nonumber
\end{equation}
where $a_i \in \mathbb{F}_2$ are the coefficients and $P(\alpha)=0$.

To better understand this field extension, compare its similarities to another
common-place
field extension $\C$, the set of complex numbers. $\C$ is an extension of the field 
of real numbers $\R$ with an additional element $i=\sqrt{-1}$, which is an imaginary
root in the algebraic closure of $\R$ -- the closure is known as the field of complex numbers $\C$.
Thus $i \notin \R$, rather $i \in \C$.
Every element $A \in \mathbb{C}$ can be represented as:
\begin{equation}\label{rep:polyC}
A=\sum_{j=0}^{1} (a_j \cdot i^j)=a_0+a_1\cdot i
\end{equation}
where $a_j \in \R$ are coefficients. Similarly, $\Fkk$ is an extension of $\F_2$ with 
an additional element $\alpha$, which is the ``imaginary root'' of an irreducible 
polynomial $P$ in $\F_2[x]$.

@@

\begin{Example}\label{exp:1}
Let us construct $\mathbb{F}_{2^4}$ as $\mathbb{F}_2[x] \pmod{ P(x)}$, where
$P(x)=x^4+x^3+1 \in \mathbb{F}_2[x]$ is an irreducible polynomial of degree $k=4$. 
Let $\alpha$ be the root of $P(x)$, {\it i.e.} $P(\alpha)=0$.

@@

\section{Normal Basis Multiplier over Galois Field}
From an algebraic perspective, a field is a space, and field elements are points in the space. Those elements can be 
represented with unique coordinates, which requires the pre-definition of a basis vector. In this
section, we discuss a special basis called normal basis, as well as the advantages adopting it in 
GF operations, especially multiplication.
\subsection{Normal Basis}
Given a Galois field $\Fkk$ is a finite field with  $2^k$ elements and characteristic equals to 2.
Its elements can be written in polynomials of $\alpha$, when there is an irreducible polynomial $p(\alpha)$
defined.

@@

\subsection{Multiplication using Normal Basis}
Besides addition/subtraction, multiplication is also very common in arithmetic circuit design.
The multiplication of GF elements in $\Fkk$ in StdB follows the rule of polynomial multiplication.
However, it will result in $O(k^2)$ bitwise operations. In other words, if we implement GF multiplication
in a bit-level logic circuit, it will contain $O(k^2)$ gates. When the datapath size $k$ is large,
the area and delay of circuit will be costly.

@@

According to the definition, a normal basis is a vector where the next entry is the square of the former one.
We note that the vector is cyclic, {\it i.e.} $\beta^{2^k} = \beta$ due to {\it Fermat's little theorem}.

@@

This example shows that convenience to use NB when computing $2^k$ power of an element.
Multiplication is more complicated than squaring; but when it is decomposed as bit-wise
operations, the property in Lemma \ref{lem:squareNB} can be well utilized.

@@

Assume the most significant bit (MSB) of the product can be represented by a function $f_{mult}$: 
\begin{equation}
\label{eqn:multiMSB}
c_{k-1} = f_{mult}(a_0, a_1, \dots, a_{k-1}; b_0, b_1, \dots, b_{k-1})
\end{equation}
Before discussing the details of the function $f_{mult}$, we can 
square both sides of Equation \ref{eqn:multiMSB}, {\it i.e.} $C^2 = A^2\times B^2$.
Obviously, using the property in Lemma \ref{lem:squareNB}, the original second most significant bit 
becomes the new MSB because of right-cyclic shifting. 
Concretely, 
$$(c_{k-1},c_0,c_1,\dots,c_{k-2}) = (a_{k-1},a_0,a_1,\dots,a_{k-2})\times(b_{k-1},b_0,b_1,\dots,b_{k-2})$$
Note $A^2, B^2$ and $C^2$ still belong to $\Fkk$, thus as a universal function implementing MSB multiplication
over $\Fkk$, $f_{mult}$ still remains the same. As a result, the new MSB can be written as 
\begin{equation}
\label{eqn:shiftMSB}
c_{k-2} = f_{mult}(a_{k-1}, a_0, a_1, 
\dots, a_{k-2}; b_{k-1}, b_0, b_1, \dots, b_{k-2})
\end{equation}
Similarly, if we take a square again on the new equation, we can get $c_{k-3}$.
Successively we can derive all bits of product $C$ using the same function $f_{mult}$, and the only
adjustment we need to make is to right-cyclic shift 2 operands by 1 bit each time.
\end{Example}

From above example, it is known that a universal structure that implements $f_{mult}$ can be reused
$k$ times in NB multiplication over $\Fkk$. Compared to StdB, which requires a distinct design 
for every bit of multiplication, NB is less costly -- as long as we can prove $f_{mult}$ implies 
a structure with $o(k^2)$ complexity (symbol $o$ denotes ``strictly lower than bound"). 
% So our next mission is to explore the details of $f_{mult}$
% to prove it will be a relatively simple design with complexity lower than $O(k^2)$.

@@

When taking different bits $l$ of the product in Equation \ref{eqn:def_lambda}, 
we obtain a series of conjugate matrices of $M$. 
Which means instead of shifting operands $A$ and $B$, we can also shift the 
matrix.

More specifically, we denote the matrix by \emph{$l$-th $\lambda$-Matrix} as 
$$c_l = A \times M^{(l)} \cdot B^T$$
Meanwhile, the operator shifting rule in Equation \ref{eqn:shiftMSB} still holds. Then we have the relation 
$$c_{l-1} = A \cdot M^{(l-1)} \cdot B^T = shift(A) \cdot M^{(l)} \cdot shift(B)^T$$ which means
by right and down cyclically shifting $M^{(l-1)}$, we can get $M^{(l)}$.

@@

If we generalize the definition and explore the nature of $\lambda$-Matrix, it is defined as cross-product terms from multiplication, which is 
\begin{equation}
Product~vector~C = (\sum_{i=0}^{k-1}a_i\beta^{2^i})(\sum_{j=0}^{k-1}b_j\beta^{2^j}) = \sum_{i=0}^{k-1}\sum_{j=0}^{k-1}a_ib_j\beta^{2^i}\beta^{2^j}
\end{equation}
The expressions $\beta^{2^i}\beta^{2^j}$ are referred to as cross-product terms, and can be represented by
NB, {\it i.e.}
\begin{equation}
\beta^{2^i}\beta^{2^j} = \sum_{l=0}^{k-1}\lambda_{ij}^{(l)}\beta^{2^l}, \ \ \lambda_{ij}^{(l)} \in \mathbb F_2.
\end{equation}
Substitution yields an expression for $l$-th digit of product as showed in Equation \ref{eqn:multiMSB}:
\begin{equation}
c_l = \sum_{i=0}^{k-1}\sum_{j=0}^{k-1}\lambda_{ij}^{(l)}a_ib_j
\end{equation}
$\lambda_{ij}^{(l)}$ is the entry with coordinate $(i,j)$ in $l$-th $\lambda$-Matrix.

The $\lambda$-Matrix can be implemented with XOR and AND gates in circuit design.
The very naive implementation requires $O(C_N)$ gates, where $C_N$ is the number of
nonzero entries in $\lambda$-Matrix.
There usually exists multiple NBs in $\Fkk, k>3$. If we employ a random NB, there is no mathematical
guarantee that $C_N$ has bound $o(k)$. However,
Mullin {\it et al.} \cite{mullinONB} % This citation is valid
proves that in certain GF $\mathbb F_{p^{k_{opt}}}$, there always exists at least one NB such that 
its corresponding $\lambda$-Matrix has $C_N = 2n-1$ nonzero entries. A basis with this property is
called optimal normal basis (ONB), details are introduced in Appendix \ref{append:ONB}.

In practice, large size NB multipliers are usually designed in $\Fkk$ when ONB exists
to minimize the number of gates. So in the following part of this chapter and our experiments,
we only focus on ONB multipliers instead of general NB multipliers.

% After fixing this, add a whole-piece StdB vs NB example, list their cost as the conclusion
\subsection{Comparison between Standard Basis and Normal Basis}
At the end of this section, a detailed example is used to make a comparison between StdB multiplication
and NB multiplication.
\begin{Example}[Rijndael's finite field]
\label{ex:Rijndael}
Rijndael uses a characteristic 2 finite field with 256 elements, which can also be called the GF $\mathbb F_{2^8}$.
Let us define the primitive element $\alpha$ using irreducible polynomial 
$\alpha^8+\alpha^7+\alpha^6+\alpha^4+\alpha^2+\alpha+1$. Coincidently, $\alpha$ is also a normal element,
i.e. $\beta = \alpha$ can construct a NB $\{\alpha,\alpha^2,\alpha^4,\alpha^8,\alpha^{16},\alpha^{32},\alpha^{64},
\alpha^{128}\}$.

@@

The final remainder is $1$, {\it i.e.} the product equals to 1 in StdB.

@@

For an 8-bit ($\mathbb F_{2^8}$) multiplier, typical design of Mastrovito multiplier consists of 218 logic gates, while 
Montgomery multiplier needs 198 gates. However, the NB multiplier reuses the $\lambda$-Matrix 
logic, so this component will only need to be implemented for once. 
Consider the definition of matrix multiplication, it needs $C_N$ AND gates to apply 
bit-wise multiplication and $C_N-1$ XOR gates to sum the intermediate products up. The number of nonzero entries
in the $\lambda$-Matrix can be counted: $C_N = 27$.
As a result, the most naive NB multiplier design (or Massey-Omura multiplier \cite{MasseyOmura})
contains 53 gates in total, which is a great saving in area cost comparing to StdB multipliers.

\section{Design of a Normal Basis Multiplier on Gate Level}
\label{sec:nbdesign}
The NB multiplier design consists of fewer gates than ordinary StdB multiplier design, even if 
we use the most naive design. However, the modern NB multiplier design has been improved a lot from the 
very first design model proposed by Massey and Omura in 1986 \cite{MasseyOmura}. In order to 
test our approach on practical contemporary circuits, it is necessary to learn the mechanism and design 
routine of several kinds of modern NB multipliers.
\subsection{Sequential Multiplier with Parallel Outputs}
The major benefit of NB multiplier origins from the sequential design. A straightforward design implementing 
the cyclic-shift of operands and $\lambda$-Matrix logic component is the Massey-Omura multiplier.

@@

The behavior of Massey-Omura multiplier can be described as: pre-load operands $A,B$ and reset $C$ to 0, after 
executing for $k$ clock cycles, the data stored in flip-flop array $C$ is the product $A\times B$.
We note that there is only one output giving 1 bit of the product each clock cycle, which matches the 
definition of serial output to communication channel. Therefore this type of design is named as 
sequential multiplier with serial output (SMSO).
The SMSO architecture need $C_N$ AND gates and $C_N-1$ XOR gates, which equals to $2k-1$ AND gates and $2k-2$ XOR gates
if it is designed using ONB. In fact, the number of gates can be reduced if the multiplication is 
implemented using a conjugate of SMSO.

@@

The gate-level logic boxes implement the following function:
\begin{equation}
\label{eqn:SMPOterms}
c_l = row_1(A\times M^{(l)}) \times B + row_2(A\times M^{(l)}) \times B + \cdots + row_{k}(A\times M^{(l)}) \times B
\end{equation}
It can be decomposed into $k$ terms. If we only compute one term for each $c_l,~0\leq l\leq k-1$ in one 
clock cycle, make $k$ outputs and add them up using shift register after $k$ clock cycles, 
it will generate the same result with SMSO. This kind of architecture is named as a 
sequential multiplier with parallel outputs (SMPO). The basic SMPO, as a conjugate of Massey-Omura multiplier,
was invented by Agnew {\it et al.} \cite{agnew1991implementation}.

@@

\begin{figure}[bp]
\centering{
\includegraphics[width=\textwidth]{newfig/mySMPO.eps}
\caption{5-bit Agnew's SMPO. Index $i$ satisfies $0<i<4$, indices $u,v$ are determined by column \# of nonzero entries in $i$-th row of $\lambda$-Matrix $M^{(0)}$, {\it i.e.} if entry $M_{ij}^{(0)}$ is a nonzero entry, $u$ or $v$ equals to $i+j \pmod 5$. Index $w = 2i\pmod 5$.}
\label{fig:SMPO}}
\end{figure}

@@

\subsection{Multiplier not based on the $\lambda$-Matrix}
Both Massey-Omura multiplier and Agnew's SMPO rely on the implementation of $\lambda$-Matrix,
which means that they will be identical if unrolled to full combinational circuits. 
After Agnew's work of parallelization, researchers proposed more designs of SMPO, 
some of them jump out of the box and are independent from $\lambda$-Matrix.
One competitive multiplier design of this type is invented by Reyhani-Masoleh and Hasan 
\cite{RHmulti}, which is therefore called RH-SMPO.

>>>>

\chapter{Finite Fields and Sequential Arithmetic Circuits}
\label{ch:prelim_GF}
This chapter provides a mathematical background for understanding 
finite fields (Galois fields) and explains how to design Galois field (GF) arithmetic circuits.
We first introduce the mathematical concepts of groups, rings, fields, and 
polynomials. 
We then apply these concepts to create Galois field arithmetic functions and 
explain how to map them to a Boolean circuit implementation.
Additionally, we introduce a special type of sequential arithmetic hardware based on normal basis, as well
as the normal basis theory behind designing such hardware.

The material is taken from \cite{galois_field:mceliece, ftheory:2006, ff:1997} for Galois field concepts, 
\cite{mastro:1989, PT:1985, acar:1998, wu:2002, Knezevic:2008} for hardware design over Galois fields, 
and previous work by Lv \cite{lv:phd}.
Normal basis theory in this section is referred from \cite{normal_book, gao:phd_normal_basis}, and sequential
normal basis arithmetic hardware designs come from \cite{mullinONB,MasseyOmura,agnew1991implementation, RHmulti}.

@@

\begin{Definition}
A {\bf field} $\mathbb{F}$ is a commutative ring with unity, where every
nonzero element in $\mathbb{F}$ has a multiplicative inverse; {\it i.e.} $\forall
a \in \mathbb{F} - \{0\}$, $\exists \hat{a} \in \mathbb{F}$ such that $ a \cdot
\hat{a} = 1$.
\end{Definition}

A field is defined as a ring with one extra condition: the presence of a 
multiplicative inverse for all nonzero elements.
Therefore, a field must be a ring while a ring is not necessarily a field.
For example, the set $\mathbb{Z}_{2^k} = \{0,1,\cdots, 2^k-1\}$ forms a finite ring.
However, $\mathbb{Z}_{2^k}$ is not a field because not every element in
$\mathbb{Z}_{2^k}$ has a multiplicative inverse. 
In the ring $\mathbb{Z}_{2^3}$, for 
instance, the element $5$ has an inverse ($5\cdot5\pmod{8}=1$) but the element $4$
does not.

An important concept in field theory is {\bf field extension}. The idea behind a
field extension is to take a base field and construct a larger field which 
contains the base field as well as satisfying additional properties. For example,
the set of real numbers $\mathbb{R}$ forms a field; one extension of 
$\mathbb{R}$ is the set of complex numbers $\mathbb{C}=\mathbb{R}(i)$. Every
element of $\mathbb{C}$ can be represented as $a+b\cdot i$ where $a,b \in \mathbb{R}$,
hence $\mathbb{C}$ is a two-dimensional extension of $\mathbb{R}$.

@@

\subsection{Finite Field}
Finite fields find widespread applications in 
many areas of electrical engineering and computer science such as error-correcting
codes, elliptic curve cryptography, digital signal processing, 
testing of VLSI circuits, among others.
In this dissertation, we specifically focus on their application to 
the FSM traversal of sequential Galois field circuits as well as abstraction refinement based on UNSAT core extraction.
This section describes the relevant Galois field concepts
\cite{galois_field:mceliece,ftheory:2006,ff:1997}
and hardware arithmetic designs over such fields \cite{mastro:1989,PT:1985,acar:1998,wu:2002,Knezevic:2008}.

@@

Algebraic extensions of the binary field $\F_{2}$  
are generally termed {\it binary extension fields} $\Fkk$.
Where elements in $\F_2$ can only represent $1$ bit, elements in $\Fkk$ 
represent a $k$-bit vector.
This allows them to be widely used in digital hardware applications.
In order to construct a Galois field of the form $\Fkk$, 
an {\bf irreducible polynomial} is required:
\begin{Definition}
A polynomial $P(x) \in \mathbb{F}_{2}\left[x\right]$, i.e., the set of all 
polynomials in $x$ with coefficients in $\F_2$, is {\bf irreducible} 
if $P(x)$ is nonconstant with degree $k$ and cannot be 
factored into a product of polynomials of lower degree in $\mathbb{F}_2[x]$.
\end{Definition}

Therefore, the polynomial $P(x)$ with degree $k$ is irreducible over 
$\mathbb{F}_{2}$ if and only if it has no roots in $\mathbb{F}_{2}$,
i.e., if $\forall a \in \mathbb{F}_{2}$, $P(a)\neq 0$.
For example, $x^2+x+1$ is an irreducible polynomial over $\mathbb{F}_{2}$
because it has no solutions in $\mathbb{F}_{2}$, i.e., $(0)^2+(0)+1=1\neq0$ 
and $(1)^2+(1)+1=1\neq0$ over $\F_2$.
Irreducible polynomials exist for any degree $\geq 2$ in $\mathbb{F}_2[x]$.

Given an irreducible polynomial $P(x)$ of degree $k$ in the polynomial ring 
$\mathbb{F}_2[x]$, we can construct a binary extension field 
$\mathbb{F}_{2^k} \equiv \mathbb{F}_2[x] \pmod{P(x)}$.
Let $\alpha$ be a root of $P(x)$, i.e., $P(\alpha)=0$.
Since $P(x)$ is irreducible over
$\mathbb{F}_2[x]$, $\alpha \notin \mathbb{F}_2$. 
Instead, $\alpha$ is an element in $\mathbb{F}_{2^k}$. 
Any element $A \in \mathbb{F}_{2^k}$ is then represented as: 
\begin{equation}\label{rep:poly}
A= \sum_{i=0}^{k-1} (a_i \cdot \alpha^i) = a_0 + a_1\cdot\alpha + \cdots + a_{k-1}\cdot \alpha^{k-1}\nonumber
\end{equation}
where $a_i \in \mathbb{F}_2$ are the coefficients and $P(\alpha)=0$.

To better understand this field extension, compare its similarities to another
commonplace field extension $\C$, the set of complex numbers. $\C$ is an extension of the field 
of real numbers $\R$ with an additional element $i=\sqrt{-1}$, which is an imaginary
root in the algebraic closure of $\R$ -- the closure is known as the field of complex numbers $\C$.
Thus $i \notin \R$, rather $i \in \C$.
Every element $A \in \mathbb{C}$ can be represented as:
\begin{equation}\label{rep:polyC}
A=\sum_{j=0}^{1} (a_j \cdot i^j)=a_0+a_1\cdot i
\end{equation}
where $a_j \in \R$ are coefficients. Similarly, $\Fkk$ is an extension of $\F_2$ with 
an additional element $\alpha$, which is the ``imaginary root'' of an irreducible 
polynomial $P$ in $\F_2[x]$.

@@

\begin{Example}\label{exp:1}
Let us construct $\mathbb{F}_{2^4}$ as $\mathbb{F}_2[x] \pmod{ P(x)}$, where
$P(x)=x^4+x^3+1 \in \mathbb{F}_2[x]$ is an irreducible polynomial of degree $k=4$. 
Let $\alpha$ be the root of $P(x)$, i.e., $P(\alpha)=0$.

@@

\section{Normal Basis Multiplier over Galois Field}
From an algebraic perspective, a field is a space, and field elements are points in the space. Those elements can be 
represented with unique coordinates, which requires the predefinition of a basis vector. In this
section, we discuss a special basis called normal basis, as well as the advantages of adopting it in 
GF operations, especially multiplication.
\subsection{Normal Basis}
Given a Galois field $\Fkk$ is a finite field with  $2^k$ elements and characteristic equals to 2.
Its elements can be written in polynomials of $\alpha$, when there is an irreducible polynomial $p(\alpha)$
defined.

@@

\subsection{Multiplication Using Normal Basis}
Besides addition/subtraction, multiplication is also very common in arithmetic circuit design.
The multiplication of GF elements in $\Fkk$ in StdB follows the rule of polynomial multiplication,
but it will result in $O(k^2)$ bitwise operations. In other words, if we implement GF multiplication
in a bit-level logic circuit, it will contain $O(k^2)$ gates. When the datapath size $k$ is large,
the area and delay of circuit will be costly.

@@

According to the definition, a normal basis is a vector where the next entry is the square of the former one.
We note that the vector is cyclic, i.e., $\beta^{2^k} = \beta$ due to {\it Fermat's little theorem}.

@@

This example shows the convenience of using NB when computing $2^k$ power of an element.
Multiplication is more complicated than squaring; but when it is decomposed as bit-wise
operations, the property in Lemma \ref{lem:squareNB} can be well utilized.

@@

Assume the most significant bit (MSB) of the product can be represented by a function $f_{mult}$: 
\begin{equation}
\label{eqn:multiMSB}
c_{k-1} = f_{mult}(a_0, a_1, \dots, a_{k-1}; b_0, b_1, \dots, b_{k-1})
\end{equation}
Before discussing the details of the function $f_{mult}$, we can 
square both sides of Equation \ref{eqn:multiMSB}, {\it i.e.}, $C^2 = A^2\times B^2$.
Obviously, using the property in Lemma \ref{lem:squareNB}, the original second most significant bit 
becomes the new MSB because of right-cyclic shifting. 
Concretely, 
$$(c_{k-1},c_0,c_1,\dots,c_{k-2}) = (a_{k-1},a_0,a_1,\dots,a_{k-2})\times(b_{k-1},b_0,b_1,\dots,b_{k-2})$$
Note $A^2, B^2$ and $C^2$ still belong to $\Fkk$, thus as a universal function implementing MSB multiplication
over $\Fkk$, $f_{mult}$ still remains the same. As a result, the new MSB can be written as 
\begin{equation}
\label{eqn:shiftMSB}
c_{k-2} = f_{mult}(a_{k-1}, a_0, a_1, 
\dots, a_{k-2}; b_{k-1}, b_0, b_1, \dots, b_{k-2})
\end{equation}
Similarly, if we take a square again on the new equation, we can get $c_{k-3}$.
Successively we can derive all bits of product $C$ using the same function $f_{mult}$, and the only
adjustment we need to make is to right-cyclic shift 2 operands by 1 bit each time.
\end{Example}

From the above example, it is known that a universal structure that implements $f_{mult}$ can be reused
$k$ times in NB multiplication over $\Fkk$. Compared to StdB, which requires a distinct design 
for every bit of multiplication, NB is less costly -- as long as we can prove $f_{mult}$ implies 
a structure with $o(k^2)$ complexity (symbol $o$ denotes ``strictly lower than bound"). 
% So our next mission is to explore the details of $f_{mult}$
% to prove it will be a relatively simple design with complexity lower than $O(k^2)$.

@@

When taking different bits $l$ of the product in Equation \ref{eqn:def_lambda}, 
we obtain a series of conjugate matrices of $M$. 
This means that instead of shifting operands $A$ and $B$, we can also shift the 
matrix.

More specifically, we denote the matrix by \emph{$l$-th $\lambda$-Matrix} as 
$$c_l = A \times M^{(l)} \cdot B^T$$
Meanwhile, the operator-shifting rule in Equation \ref{eqn:shiftMSB} still holds. Then we have the relation 
$$c_{l-1} = A \cdot M^{(l-1)} \cdot B^T = shift(A) \cdot M^{(l)} \cdot shift(B)^T$$ which means
by right and down cyclically shifting $M^{(l-1)}$, we can get $M^{(l)}$.

@@

If we generalize the definition and explore the nature of $\lambda$-Matrix, it is defined as cross-product terms from multiplication, which is 
\begin{equation}
Product~vector~C = (\sum_{i=0}^{k-1}a_i\beta^{2^i})(\sum_{j=0}^{k-1}b_j\beta^{2^j}) = \sum_{i=0}^{k-1}\sum_{j=0}^{k-1}a_ib_j\beta^{2^i}\beta^{2^j}
\end{equation}
The expressions $\beta^{2^i}\beta^{2^j}$ are referred to as cross-product terms, and can be represented by
NB, i.e.,
\begin{equation}
\beta^{2^i}\beta^{2^j} = \sum_{l=0}^{k-1}\lambda_{ij}^{(l)}\beta^{2^l}, \ \ \lambda_{ij}^{(l)} \in \mathbb F_2.
\end{equation}
Substitution yields an expression for $l$-th digit of product as showed in Equation \ref{eqn:multiMSB}:
\begin{equation}
c_l = \sum_{i=0}^{k-1}\sum_{j=0}^{k-1}\lambda_{ij}^{(l)}a_ib_j
\end{equation}
$\lambda_{ij}^{(l)}$ is the entry with coordinate $(i,j)$ in $l$-th $\lambda$-Matrix.

The $\lambda$-Matrix can be implemented with XOR and AND gates in circuit design.
The very naive implementation requires $O(C_N)$ gates, where $C_N$ is the number of
nonzero entries in $\lambda$-Matrix.
There usually exist multiple NBs in $\Fkk, k>3$. If we employ a random NB, there is no mathematical
guarantee that $C_N$ has bound $o(k)$. However,
Mullin {\it et al.} \cite{mullinONB} % This citation is valid
prove that in certain GF $\mathbb F_{p^{k_{opt}}}$, there always exists at least one NB such that 
its corresponding $\lambda$-Matrix has $C_N = 2n-1$ nonzero entries. A basis with this property is
called optimal normal basis (ONB), details of which are introduced in Appendix \ref{append:ONB}.

In practice, large NB multipliers are usually designed in $\Fkk$ when ONB exists
to minimize the number of gates. So in the following part of this chapter and our experiments,
we only focus on ONB multipliers instead of general NB multipliers.

% After fixing this, add a whole-piece StdB vs NB example, list their cost as the conclusion
\subsection{Comparison between Standard Basis and Normal Basis}
At the end of this section, a detailed example is used to make a comparison between StdB multiplication
and NB multiplication.
\begin{Example}[Rijndael's finite field]
\label{ex:Rijndael}
Rijndael uses a characteristic 2 finite field with 256 elements, which can also be called the GF $\mathbb F_{2^8}$.
Let us define the primitive element $\alpha$ using irreducible polynomial 
$\alpha^8+\alpha^7+\alpha^6+\alpha^4+\alpha^2+\alpha+1$. Coincidently, $\alpha$ is also a normal element,
i.e., $\beta = \alpha$ can construct a NB $\{\alpha,\alpha^2,\alpha^4,\alpha^8,\alpha^{16},\alpha^{32},\alpha^{64},
\alpha^{128}\}$.

@@

The final remainder is $1$, {\it i.e.} the product equals 1 in StdB.

@@

For an 8-bit ($\mathbb F_{2^8}$) multiplier, typical design of Mastrovito multiplier consists of 218 logic gates, while 
Montgomery multiplier needs 198 gates. However, the NB multiplier reuses the $\lambda$-Matrix 
logic, so this component will only need to be implemented for once. 
Consider the definition of matrix multiplication, it needs $C_N$ AND gates to apply 
bit-wise multiplication and $C_N-1$ XOR gates to sum the intermediate products up. The number of nonzero entries
in the $\lambda$-Matrix can be counted: $C_N = 27$.
As a result, the most naive NB multiplier design (or Massey-Omura multiplier \cite{MasseyOmura})
contains 53 gates in total, which is a great saving in area cost compared to StdB multipliers.

\section{Design of a Normal Basis Multiplier on Gate Level}
\label{sec:nbdesign}
The NB multiplier design consists of fewer gates than ordinary StdB multiplier design, even if 
we use the most naive design. However, the modern NB multiplier design has been improved a lot from the 
very first design model proposed by Massey and Omura in 1986 \cite{MasseyOmura}. In order to 
test our approach on practical contemporary circuits, it is necessary to learn the mechanism and design 
routine of several kinds of modern NB multipliers.
\subsection{Sequential Multiplier with Parallel Outputs}
The major benefit of NB multiplier originates from the sequential design. A straightforward design implementing 
the cyclic-shift of operands and $\lambda$-Matrix logic component is the Massey-Omura multiplier.

@@

The behavior of Massey-Omura multiplier can be described as follows: preload operands $A,B$ and reset $C$ to 0, after 
executing for $k$ clock cycles, the data stored in flip-flop array $C$ is the product $A\times B$.
We note that there is only one output giving 1 bit of the product each clock cycle, which matches the 
definition of serial output to communication channel. Therefore this type of design is called a 
sequential multiplier with serial output (SMSO).
The SMSO architecture needs $C_N$ AND gates and $C_N-1$ XOR gates, which equals to $2k-1$ AND gates and $2k-2$ XOR gates
if it is designed using ONB. In fact, the number of gates can be reduced if the multiplication is 
implemented using a conjugate of SMSO.

@@

The gate-level logic boxes implement the following function:
\begin{equation}
\label{eqn:SMPOterms}
c_l = row_1(A\times M^{(l)}) \times B + row_2(A\times M^{(l)}) \times B + \cdots + row_{k}(A\times M^{(l)}) \times B
\end{equation}
It can be decomposed into $k$ terms. If we only compute one term for each $c_l,~0\leq l\leq k-1$ in one 
clock cycle, make $k$ outputs and add them up using shift register after $k$ clock cycles, 
it will generate the same result with SMSO. This kind of architecture is called a 
sequential multiplier with parallel outputs (SMPO). The basic SMPO, as a conjugate of Massey-Omura multiplier,
was invented by Agnew {\it et al.} \cite{agnew1991implementation}.

@@

\begin{figure}[bp]
\centering{
\includegraphics[width=\textwidth]{newfig/mySMPO.eps}
\caption{5-bit Agnew's SMPO. Index $i$ satisfies $0<i<4$, indices $u,v$ are determined by column \# of nonzero entries in $i$-th row of $\lambda$-Matrix $M^{(0)}$, i.e., if entry $M_{ij}^{(0)}$ is a nonzero entry, $u$ or $v$ equals to $i+j \pmod 5$. Index $w = 2i\pmod 5$.}
\label{fig:SMPO}}
\end{figure}

@@

\subsection{Multiplier Not Based on the $\lambda$-Matrix}
Both Massey-Omura multiplier and Agnew's SMPO rely on the implementation of $\lambda$-Matrix,
which means that they will be identical if unrolled to full combinational circuits. 
After Agnew's work of parallelization, researchers proposed more designs of SMPO, 
some of which jump out of the box and are independent from $\lambda$-Matrix.
One competitive multiplier design of this type was invented by Reyhani-Masoleh and Hasan 
\cite{RHmulti}, which is therefore called RH-SMPO.

Figure \ref{fig:5bitRH} is a 5-bit RH-SMPO, which is functionally equivalent to the 5-bit Agnew's SMPO 
in Figure \ref{fig:SMPO}. A brief proof is as follows:
